SmartCheck: Automatic and Efficient Counterexample Reduction and Generalization\\$\;$\\DRAFT

### 2013-03-03 paper/paper.tex
paper updates.

\title{SmartCheck: Automatic and Efficient Counterexample Reduction and Generalization}
%\subtitle{Subtitle Text, if any}

@@

\noindent
In Section~\ref{sec:experiments}, we discuss some of our experiences with using
SmartCheck, including checking properties from Xmonad, and related work and
concluding remarks are made in Sections~\ref{sec:related}
and~\ref{sec:conclusions}, respectively.

@@

\section{Experiments}\label{sec:experiments}
Here we describe three experiments using SmartCheck, including an Xmonad
property and a property about a natural language processing library.

@@

\subsection{Xmonad}
Xmonad is a popular X11 window manager written in Haskell~\cite{xmonad}.  Xmonad
is a large software project with many contributors, so consequently, a
QuickCheck test harness is included to help ensure new commits do not introduce
bugs.  At the heart of Xmonad is a \ttp{StackSet} data type that encodes the
relationship between windows, work spaces, and which window has the focus.
Xmonad contains properties to ensure the correct manipulation of
\ttp{StackSet}s.

Xmonad passes all of its QuickCheck tests, but let us see what might happen to a
new contributor if things go awry.  Suppose a developer defines an insertion
function that inserts a new focus into the \ttp{StackSet}, except she forgets to
place the value inside.  The correct insertion function from Xmonad is
%
\begin{code}
insertUp a s = if member a s then s else insert
  where
  insert = modify (Just $ Stack a [] [])
                  (\textbackslash(Stack t l r) ->
                      Just $ Stack a l (t:r)) s
\end{code}
%
\noindent
So suppose the developer forgets to cons \ttp{t} onto \ttp{r}.  Fortunately,
there are QuickCheck tests that depend on \ttp{insertUp}.  One such test is
\ttp{prop\_shift\_reversible}, which tests if the \ttp{shift} function, which
checks a property about shifting focus back and forth between windows.

With the incorrect version of \ttp{insertUp}, an example returned by QuickCheck
would typically include a \ttp{StackSet} value as follows:

\medskip%
\begin{sloppypar}
\scriptsize
\noindent%
\ttp{StackSet \{current = Screen \{workspace = Workspace \{tag = NonNegative
  \{getNonNegative = 5\}, layout = -2, stack = Just (Stack \{focus = 'I', up =
  ", down = "\})\}, screen = 5, screenDetail = 0\}, visible = [Screen
    \{workspace = Workspace \{tag = NonNegative \{getNonNegative = 0\}, layout =
    -2, stack = Just (Stack \{focus = `\char`\\161', up = ", down =
    "\})\}, screen = 1, screenDetail = -2\},Screen \{workspace = Workspace
    \{tag = NonNegative \{getNonNegative = 2\}, layout = -2, stack = Just (Stack
    \{focus = '3', up = ", down = "\})\}, screen = 2, screenDetail =
    -1\},Screen \{workspace = Workspace \{tag = NonNegative \{getNonNegative =
    3\}, layout = -2, stack = Just (Stack \{focus = 'T', up = ", down =
    "\})\}, screen = 3, screenDetail ) = 2\},Screen \{workspace = Workspace
    \{tag = NonNegative \{getNonNegative = 4\}, layout = -2, stack = Nothing\},
    screen = 4, screenDetail = -1\},Screen \{workspace = Workspace \{tag =
    NonNegative \{getNonNegative = 6\}, layout = -2, stack = Just (Stack \{focus
    = 'm', up = ", down = "\})\}, screen = 6, screenDetail = 2\},Screen
    \{workspace = Workspace \{tag = NonNegative \{getNonNegative = 7\}, layout =
    -2, stack = Nothing\}, screen = 0, screenDetail = 0\}], hidden = [Workspace
    \{tag = NonNegative \{getNonNegative = 1\}, layout = -2, stack = Just (Stack
    \{focus = `\char`\\DC3', up = ", down = "\})\},Workspace \{tag = NonNegative \{getNonNegative = 8\}, layout = -2, stack = Nothing\},Workspace \{tag = NonNegative \{getNonNegative = 9\}, layout = -2, stack = Just (Stack \{focus = `\char`\\ACK', up = ", down = "\})\}], floating = fromList []\} }
\end{sloppypar}
\medskip%
\noindent
This is a wall of text, which provides little useful information about what went
wrong.  SmartCheck, however, takes the value and returns the following:
%
\begin{code}
Extrapolated value:
forall x0 x1 x2:
  StackSet x0 x1 x2 (fromList [])
\end{code}
%
\noindent
\todo{double check}
The programmer immediately sees that the first three fields in the record are
irrelevant.  The upshot is that any \ttp{StackSet} fails because insertion is
ill-defined.  Such an insight, while obvious in hind-sight, can take hours to
arrive at as the programmer manually tries to write her own shrink functions or
manually deconstruct a large record.

\subsection{Natural Language Processing}
In 2012, a question was posted on the programming message board
\emph{stackoverflow} asking about how to shrinking large nested data
types.\footnote{\url{http://stackoverflow.com/questions/8788542/how-do-i-get-good-small-shrinks-out-of-quickcheck}}
The poster writes:
\begin{quote}
$\ldots$ I tend to get an incomprehensible page full of output. $\ldots$
  Implementing the shrink function for each of my types seems to help a little,
  but not as much as I'd like.  $\ldots$ If I try to tune my shrink
  implementations, I also find that QC starts taking a very long time.
\end{quote}
\noindent
The question relates to the Geni natural language processing (NLP) package
implemented in Haskell~\cite{kow}.  Specifically, counterexamples to a property
attempting to show that a macro expansion function is its own inverse are
enormous, requiring between 150 to 200 console output lines. \todo{redo tests}

Using SmartCheck, we are able to reduce counterexamples to around 25 lines of
output.  Most of the savings in the counterexample size were due to universal
generalization, like in the Xmonad case: entire record fields are abstracted
away.

``Retro-fitting'' SmartCheck to GenI, which is a large software system, requires
some work.  Fortunately, because the program already uses QuickCheck,
\ttp{arbitrary} instances are already defined, but we had to ensure all relevant
types derived the \ttp{Read}, \ttp{Show}, \ttp{Typeable}, and \ttp{Generic}
type classes as well, together with \ttp{SubTypes}.  The work does not require
understanding the program, but it does require hunting for definitions and
multiple rounds of recompilation.

>>>>

\title{SmartCheck: Automatic and Efficient Counterexample Reduction and Generalization\\$\;$\\DRAFT}
%\subtitle{Subtitle Text, if any}

@@

\noindent
In Section~\ref{sec:experiments}, we discuss some of our experiences with using
SmartCheck, including checking properties from XMonad, and related work and
concluding remarks are made in Sections~\ref{sec:related}
and~\ref{sec:conclusions}, respectively.

@@

\section{Experiments}\label{sec:experiments}
Here we describe three experiments using SmartCheck, including an XMonad
property and a property about a natural language processing library.

@@

\subsection{XMonad}
XMonad is a popular X11 window manager written in Haskell~\cite{xmonad}.  XMonad
is a large software project with many contributors, so consequently, a
QuickCheck test harness is included to help ensure new commits do not introduce
bugs.  At the heart of XMonad is a \ttp{StackSet} data type that encodes the
relationship between windows, work spaces, and which window has the focus.
XMonad contains properties to ensure the correct manipulation of
\ttp{StackSet}s.

XMonad passes all of its QuickCheck tests, but let us see what might happen to a
new contributor if things go awry.  Suppose a developer defines a deletion
function to delete a window, if it exists.  There is a deletion function that
exists in XMonad, which is quite complex, given the amount of state that is managed
by \ttp{StackSet}.  However, one function used in deletion is to filter the
stack of windows associated with each workspace defined:
%
\begin{code}
removeFromWorkspace ws =
  ws { stack = stack ws >>= filter (/=w) }
\end{code}
%
\noindent
Now, suppose the programmer makes a simple typo (or perhaps has a more serious
misunderstanding) and instead writes
%
\begin{code}
removeFromWorkspace ws =
  ws { stack = stack ws >>= filter (==w) }
\end{code}
%
\noindent
When testing the property \ttp{prop\_delete}, which says that deleting the
focused window of the current stack removes it from the \ttp{StackSet} \ttp{x}.
%
\begin{code}
prop_delete x =
    case peek x of
        Nothing -> True
        Just i  -> not (member i (delete i x))
\end{code}
%
\noindent
QuickCheck returns
%
\medskip%
\begin{sloppypar}
\scriptsize
\noindent%
\ttp{StackSet \{current = Screen \{workspace = Workspace \{tag = NonNegative
      \{getNonNegative = 0\}, layout = -1, stack = Just (Stack \{focus = 'S', up =
        "", down = ""\})\}, screen = 1, screenDetail = 1\}, visible = [Screen
    \{workspace = Workspace \{tag = NonNegative \{getNonNegative = 2\}, layout = -1,
        stack = Nothing\}, screen = 2, screenDetail = -1\},Screen \{workspace =
      Workspace \{tag = NonNegative \{getNonNegative = 3\}, layout = -1, stack =
        Nothing\}, screen = 0, screenDetail = -1\}], hidden = [Workspace \{tag =
      NonNegative \{getNonNegative = 1\}, layout = -1, stack = Just (Stack \{focus
        = `\char`\\NUL', up = "", down = ""\})\},Workspace \{tag = NonNegative
      \{getNonNegative = 4\}, layout = -1, stack = Just (Stack \{focus = 'I', up =
        "", down = ""\})\}], floating = fromList []\} }
\end{sloppypar}
\medskip%
\noindent
This is a relatively small counterexample, but even the smallest \ttp{StackSet}
values are somewhat large due to the number of fields within it.  But let us see
if we can do better.

SmartCheck returns a smaller counterexample, and then generalizes it:
%
\begin{code}
forall values x0 x1 x2 x3:
  StackSet
    (Screen (Workspace x0 (-1) (Just x1)) 1 1)
    x2 x3 (fromList [])
\end{code}
%
\noindent
We can look at what was generalized and what was not.  First, the last element
of \ttp{StackSet} is a map, and we treated it as a base type, so it is not
analyzed at all.  But it is an empty map.  The second and third fields of
\ttp{StackSet} are generalized, but the first one is not.  So there is something
particular about it.  So the culprit is one of the small constants (1 and -1) or
having a \ttp{Just} value rather than a \ttp{Nothing}: it turns out that what
matters is having a \ttp{Just} value, which is the stack field that deletion
works on.

%% n insertion
%% function that inserts a new focus into the \ttp{StackSet}, except she forgets to
%% place the value inside.  The correct insertion function from XMonad is
%% %
%% \begin{code}
%% insertUp a s = if member a s then s else insert
%%   where
%%   insert = modify (Just $ Stack a [] [])
%%                   (\textbackslash(Stack t l r) ->
%%                       Just $ Stack a l (t:r)) s
%% \end{code}
%% %
%% \noindent
%% So suppose the developer forgets to cons \ttp{t} onto \ttp{r}.  Fortunately,
%% there are QuickCheck tests that depend on \ttp{insertUp}.  One such test is
%% \ttp{prop\_shift\_reversible}, which tests if the \ttp{shift} function, which
%% checks a property about shifting focus back and forth between windows.

%% With the incorrect version of \ttp{insertUp}, an example returned by QuickCheck
%% would typically include a \ttp{StackSet} value as follows:

%% \medskip%
%% \begin{sloppypar}
%% \scriptsize
%% \noindent%
%% \ttp{StackSet \{current = Screen \{workspace = Workspace \{tag = NonNegative
%%   \{getNonNegative = 5\}, layout = -2, stack = Just (Stack \{focus = 'I', up =
%%   ", down = "\})\}, screen = 5, screenDetail = 0\}, visible = [Screen
%%     \{workspace = Workspace \{tag = NonNegative \{getNonNegative = 0\}, layout =
%%     -2, stack = Just (Stack \{focus = `\char`\\161', up = ", down =
%%     "\})\}, screen = 1, screenDetail = -2\},Screen \{workspace = Workspace
%%     \{tag = NonNegative \{getNonNegative = 2\}, layout = -2, stack = Just (Stack
%%     \{focus = '3', up = ", down = "\})\}, screen = 2, screenDetail =
%%     -1\},Screen \{workspace = Workspace \{tag = NonNegative \{getNonNegative =
%%     3\}, layout = -2, stack = Just (Stack \{focus = 'T', up = ", down =
%%     "\})\}, screen = 3, screenDetail ) = 2\},Screen \{workspace = Workspace
%%     \{tag = NonNegative \{getNonNegative = 4\}, layout = -2, stack = Nothing\},
%%     screen = 4, screenDetail = -1\},Screen \{workspace = Workspace \{tag =
%%     NonNegative \{getNonNegative = 6\}, layout = -2, stack = Just (Stack \{focus
%%     = 'm', up = ", down = "\})\}, screen = 6, screenDetail = 2\},Screen
%%     \{workspace = Workspace \{tag = NonNegative \{getNonNegative = 7\}, layout =
%%     -2, stack = Nothing\}, screen = 0, screenDetail = 0\}], hidden = [Workspace
%%     \{tag = NonNegative \{getNonNegative = 1\}, layout = -2, stack = Just (Stack
%%     \{focus = `\char`\\DC3', up = ", down = "\})\},Workspace \{tag = NonNegative \{getNonNegative = 8\}, layout = -2, stack = Nothing\},Workspace \{tag = NonNegative \{getNonNegative = 9\}, layout = -2, stack = Just (Stack \{focus = `\char`\\ACK', up = ", down = "\})\}], floating = fromList []\} }
%% \end{sloppypar}
%% \medskip%
%% \noindent
%% This is a wall of text, which provides little useful information about what went
%% wrong.  SmartCheck, however, takes the value and returns the following:
%% %
%% \begin{code}
%% Extrapolated value:
%% forall values x0 x1 x2:
%%   StackSet x0 x1 x2 (fromList [])
%% \end{code}
%% %
%% \noindent
%% (The last field of \ttp{StackSet} is a map, and we treat it as a base type.)
%% The programmer immediately sees that failing instances were found when
%% substituting arbitrary well-typed values into the first three fields of
%% \ttp{StackSet}.  Therefore, there is likely nothing specific in the
%% counterexample causing the failures.

%% the first three fields in the record are
%% irrelevant.  The upshot is that any \ttp{StackSet} fails because insertion is
%% ill-defined.  Such an insight, while obvious in hind-sight, can take hours to
%% arrive at as the programmer manually tries to write her own shrink functions or
%% manually deconstruct a large record.

\subsection{Natural Language Processing}
In 2012, a question was posted on the programming message board
Stack Overflow asking about how to shrinking large nested data
types.\footnote{\url{http://stackoverflow.com/questions/8788542/how-do-i-get-good-small-shrinks-out-of-quickcheck}}
The poster writes:
\begin{quote}
$\ldots$ I tend to get an incomprehensible page full of output. $\ldots$
  Implementing the shrink function for each of my types seems to help a little,
  but not as much as I'd like.  $\ldots$ If I try to tune my shrink
  implementations, I also find that QC starts taking a very long time.
\end{quote}
\noindent
The question relates to the Geni natural language processing (NLP) package
implemented in Haskell~\cite{kow}.  Specifically, counterexamples to a property
attempting to show that a macro expansion function is its own inverse are
enormous, requiring 200-300 80-character lines to print.

Using SmartCheck, we are able to reduce counterexamples to around 25
80-character lines of output.  Most of the savings in the counterexample size
were due to universal generalization, like in the XMonad case: entire record
fields are abstracted away.

``Retro-fitting'' SmartCheck to GenI, which is a large software system, required
some work.  Fortunately, because the program already uses QuickCheck,
\ttp{arbitrary} instances are already defined, but we had to ensure all relevant
types derived the \ttp{Read}, \ttp{Show}, \ttp{Typeable}, and \ttp{Generic}
type classes as well, together with \ttp{SubTypes}.  The work does not require
understanding the program, but it does require hunting for definitions and
multiple rounds of recompilation.
